Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-28 paper.tex
Pass through AoU data

\subsection{Case study: All of Us exome-like data}
The All of Us dataset comprises 245K clinical-grade whole genome sequences,
identifying over 1 billion variants, including 3.9 million novel variants
with coding consequences~\citep{all2024genomic}. Participants are from the United States,
with an emphasis on historically underrepresented genetic ancestry groups.
The genomic data, generated through short-read sequencing on the Illumina NovaSeq 6000,
is of high quality, with an average coverage depth of 30x, and is linked to extensive
health-related information, providing a valuable resource for biomedical research.
WGS analysis is conducted on the Illumina DRAGEN platform, with alignment performed
using the GRCh38dh reference genome. GTC files are converted to VCF format using
Picard GtcToVcf, and every step of the WGS procedure is rigorously controlled by
predefined quality control measures.

For this study, we converted the chromosome 20 exome-like
data from the All of Us dataset to VCF Zarr.
The exome-like data is constructed by extracting from all WGS samples
the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic
transcripts. This dataset, stored as a single compressed VCF file of 7.44 GiB,
includes 715,256 variants, of which more than 700 are classified as pathogenic or likely pathogenic.

@@

[TODO discuss local alleles and mention Table~\ref{tab-aou-data}].

[Discuss why we're bigger than VCF here, people will want to understand.

>>>>

\subsection{Case study: All of Us exome-like data}
The All of Us dataset comprises 245K clinical-grade whole genome sequences,
identifying over 1 billion variants, including 3.9 million novel variants
with coding consequences~\citep{all2024genomic}. 
Participants are from the United States,
with an emphasis on historically underrepresented genetic ancestry groups.
The genomic data, generated through short-read sequencing on the Illumina NovaSeq 6000,
is of high quality, with an average coverage depth of 30x, and is linked to extensive
health-related information.
For this study, we converted the chromosome 20 exome-like
data from the All of Us dataset to VCF Zarr.
The exome-like data is constructed by extracting from all WGS samples
the SNP and indel variants that lie within the 
chromosome 20 exon regions of the Gencode v42 basic
transcripts. This dataset, stored as a single compressed VCF file of 7.44 GiB,
includes 715,256 variants, of which more than 700 are classified as 
pathogenic or likely pathogenic.

@@

Local alleles (see Methods) is essential in this example
as the maximum number of alleles observed at a site is 95.
Thus, as defined, the AD field field has an inner dimension of 95
each $1000\times10000$ chunk 
% >>> 1000 * 10_000 * 95 * 2 / 1024**3
% 1.7695128917694092
has an uncompressed size of 1.77GiB (in this 16 bit integers are required
to store the values). Given that the vast majority
of this encoding consists of the dimension padding sentinel (-2, see 
Methods), it is an inefficient means of representing and processing 
allele depth data. Translating to the local alleles representation
here reduces the inner dimension to 2, and the uncompressed 
% >>> 1000 * 10_000 * 2 * 2 / 1024**2
% 38.14697265625
size of each chunk to 38.15MiB. 
Note that there is also no loss of information in this case 
by encoding as local alleles, as the per-sample gVCFs used 
in earlier stages in the All of Us pipeline were truncated
to a maximum of two ALT alleles per site.

[Discussion of the sparsity of the QC field data, and how this 
explains the difference to the VCF compression.]
